Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-12 talk/pepm2011/paper.tex
We plan to have only one conclusion, not many.
clean up figures
actually reference this figure
(cfbolz, arigo) Conclusion.

For the purpose of this paper, we are going to use a very simple object
model, that just supports an integer and a float type. The objects support only
two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
than zero. The implementation of \texttt{add} uses classical Smalltalk-like
double-dispatching. These classes could be part of the implementation of a very
simple interpreter written in RPython.

@@

class BoxedFloat(Base):
    def __init__(self, floatval):
        self.floatval = floatval
    def add(self, other):
        return other.add__float(self.floatval)
    def add__int(self, intother):
        floatvalue = float(intother) + self.floatval
        return BoxedFloat(floatvalue)
    def add__float(self, floatother):
        return BoxedFloat(floatother + self.floatval)
    def is_positive(self):
        return self.floatval > 0.0
\end{verbatim}
\caption{A simple object model}
\end{figure}

@@

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\label{fig:unopt-trace}
\caption{Unoptimized Trace for the Simple Object Model}
\end{figure}

@@

\begin{figure}
\includegraphics{figures/step1.pdf}
\label{fig:step1}
\caption{Resulting Trace After Allocation Removal}
\end{figure}

The final optimized trace of the example can be seen in
Figure~\ref{fig:step1}.
XXX why does it says ``Figure 4.1'' here but ``Figure 4'' in the label?

The optimized trace contains only two allocations, instead of the original five,
and only three \texttt{guard\_class} operations, from the original seven.

@@

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\label{fig:semantics}
\caption{The Operational Semantics of Simplified Traces}
\end{figure*}

@@

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
    u^*,v^*,w^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Static\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\label{fig:optimization}
\caption{Optimization Rules}
\end{figure*}

To optimize the simple traces from the last section, we use online partial
evaluation. The partial evaluator optimizes one operation of the trace at a
time. Every operation in the unoptimized trace is replaced by a list of
operations in the optimized trace. This list is empty if the operation
can be optimized away (which hopefully happens often).

@@

\section{Conclusions}
\label{sec:conclusions}

>>>>

For the purpose of this paper, we are going to use a very simple object
model, that just supports an integer and a float type. The objects support only
two operations, \texttt{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \texttt{is\_positive}, which returns whether the number is greater
than zero. The implementation of \texttt{add} uses classical Smalltalk-like
double-dispatching. These classes could be part of the implementation of a very
simple interpreter written in RPython. The classes can be seen in
Figure~\ref{fig:objmodel}.

@@

class BoxedFloat(Base):
    def __init__(self, floatval):
        self.floatval = floatval
    def add(self, other):
        return other.add__float(self.floatval)
    def add__int(self, intother):
        floatvalue = float(intother) + self.floatval
        return BoxedFloat(floatvalue)
    def add__float(self, floatother):
        return BoxedFloat(floatother + self.floatval)
    def is_positive(self):
        return self.floatval > 0.0
\end{verbatim}
\caption{A simple object model}
\label{fig:objmodel}
\end{figure}

@@

\begin{figure}
\texttt{
\begin{tabular}{l} 
\# arguments to the trace: $p_{0}$, $p_{1}$ \\
\# inside f: res.add(y) \\
guard\_class($p_{1}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{2}$ = get($p_{1}$, intval) \\
~~~~guard\_class($p_{0}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{3}$ = get($p_{0}$, intval) \\
~~~~~~~~$i_{4}$ = int\_add($i_{2}$, $i_{3}$) \\
~~~~~~~~$p_{5}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{5}$, intval, $i_{4}$) \\
\# inside f: BoxedInteger(-100)  \\
$p_{6}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{6}$, intval, -100) \\
~\\
\# inside f: .add(BoxedInteger(-100)) \\
guard\_class($p_{5}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{7}$ = get($p_{5}$, intval) \\
~~~~guard\_class($p_{6}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{8}$ = get($p_{6}$, intval) \\
~~~~~~~~$i_{9}$ = int\_add($i_{7}$, $i_{8}$) \\
~~~~~~~~$p_{10}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{10}$, intval, $i_{9}$) \\
~\\
\# inside f: BoxedInteger(-1) \\
$p_{11}$ = new(BoxedInteger) \\
~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~set($p_{11}$, intval, -1) \\
~\\
\# inside f: y.add(BoxedInteger(-1)) \\
guard\_class($p_{0}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.add \\
~~~~$i_{12}$ = get($p_{0}$, intval) \\
~~~~guard\_class($p_{11}$, BoxedInteger) \\
~~~~~~~~\# inside BoxedInteger.add\_\_int \\
~~~~~~~~$i_{13}$ = get($p_{11}$, intval) \\
~~~~~~~~$i_{14}$ = int\_add($i_{12}$, $i_{13}$) \\
~~~~~~~~$p_{15}$ = new(BoxedInteger) \\
~~~~~~~~~~~~\# inside BoxedInteger.\_\_init\_\_ \\
~~~~~~~~~~~~set($p_{15}$, intval, $i_{14}$) \\
~\\
\# inside f: y.is\_positive() \\
guard\_class($p_{15}$, BoxedInteger) \\
~~~~\# inside BoxedInteger.is\_positive \\
~~~~$i_{16}$ = get($p_{15}$, intval) \\
~~~~$i_{17}$ = int\_gt($i_{16}$, 0) \\
\# inside f \\
guard\_true($i_{17}$) \\
jump($p_{15}$, $p_{10}$) \\
\end{tabular}
}
\caption{Unoptimized Trace for the Simple Object Model}
\label{fig:unopt-trace}
\end{figure}

@@

\begin{figure}
\includegraphics{figures/step1.pdf}
\caption{Resulting Trace After Allocation Removal}
\label{fig:step1}
\end{figure}

The final optimized trace of the example can be seen in Figure~\ref{fig:step1}.
The optimized trace contains only two allocations, instead of the original five,
and only three \texttt{guard\_class} operations, from the original seven.

@@

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v & \in & V & \mathrm{\ variables\ in\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
    l & \in & L & \mathrm{\ locations\ on\ heap}
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup L & \mathrm{\ Environment}\\
    H & \in & L\rightharpoonup\mathfrak{T}\times(L\cup \{ \mathrm{null} \})\times (L\cup \{ \mathrm{null}\} ) & \mathrm{\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\caption{The Operational Semantics of Simplified Traces}
\label{fig:semantics}
\end{figure*}

@@

\begin{minipage}[b]{7 cm}
\emph{Object Domains:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    u,v,w & \in & V & \mathrm{\ variables\ in\ trace}\\
    u^*,v^*,w^* & \in & V^* & \mathrm{\ variables\ in\ optimized\ trace}\\
    T & \in & \mathfrak{T} & \mathrm{\ runtime\ types}\\
    F & \in & \left\{ L,R\right\} & \mathrm{\ fields\ of\ objects}\\
 \end{array}
$$
\end{minipage}
\begin{minipage}[b]{5 cm}
\emph{Semantic Values:}
$$\setlength\arraycolsep{0.1em}
 \begin{array}{rcll}
    E & \in & V\rightharpoonup V^* & \mathrm{\ Environment}\\
    S & \in & V^*\rightharpoonup\mathfrak{T}\times(V^*\cup \{ \mathrm{null} \})\times (V^*\cup \{ \mathrm{null}\} ) & \mathrm{\ Static\ Heap}\\
    \\
    \\
 \end{array}
$$
\end{minipage}
\end{center}
\caption{Optimization Rules}
\label{fig:optimization}
\end{figure*}

To optimize the simple traces from the last section, we use online partial
evaluation. The partial evaluator optimizes one operation of the trace at a
time. Every operation in the unoptimized trace is replaced by a list of
operations in the optimized trace. This list is empty if the operation
can be optimized away (which hopefully happens often). The optimization rules
can be seen in Figure~\ref{fig:optimization}.

@@

\section{Conclusion}
\label{sec:conclusion}

In this paper, we used a approach based on partial evaluation to
optimize allocations in the traces of a tracing JIT.  In this context a
simple approach to partial evaluation gives good results.  This is due
to the fact that the tracing JIT itself is responsible for all control
issues, which are usually the hardest part of partial evaluation: the
tracing JIT selects the parts of the program that are worthwhile to
optimize, and extracts linear paths through them, inlining functions as
necessary.  What is left to optimize is only those linear paths.

We expect a similar result for other optimizations that usually require
a complex analysis phase and are thus normally too slow to use at
runtime.  A tracing JIT selects interesting linear paths by itself;
therefore, a naive version of many optimizations on such paths should
give mostly the same results.  For example, we experimented with (and
plan to write about) store-load propagation with a very simple alias
analysis.
